Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-04-05 ms_pre_post.Rmd
Minor tweaks to Methods

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because of the possibility for inbreeding preference to evolve in the model), by tracking the dynamics of alleles underlying reproductive strategies enacted by interacting females and males within a small focal population. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection. Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of the three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$), pre-copulatory inbreeding strategy ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each of the three traits whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. In overview, each generation proceeds with females paying costs and expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values for tendency for polyandry ($P_{p}$), pre-copulatory inbreeding strategy ($M_{p}$), and post-copulatory inbreeding strategy ($F_{p}$) each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively (see details of mating and fertilisation strategies below). In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see details of polyandry below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of her offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. We simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are permitted to evolve, and could potentially arise due to mutation or drift.

@@

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ did not evolve). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.

>>>>

We model evolution of polyandry, and of pre-copulatory and post-copulatory inbreeding avoidance strategies (hereafter simply 'inbreeding strategies' because the model did not preclude evolution of inbreeding preference), by tracking the dynamics of alleles underlying reproductive strategies enacted by interacting females and males within a small focal population. We thereby track evolutionary dynamics given internally consistent patterns of relatedness caused by non-random mating and allowing for effects of mutation, gene flow, drift, and selection. Each individual has 10 diploid loci, and therefore 20 alleles, underlying each of three reproductive strategy traits (i.e., 30 diploid loci in total): tendency for polyandry ($P_{a}$, i.e., 'polyandry' alleles), pre-copulatory inbreeding strategy ($M_{a}$, i.e., 'mating' alleles), and post-copulatory inbreeding strategy ($F_{a}$, i.e., 'fertilisation' alleles). All alleles can take the value of any real number [i.e., a continuum-of-alleles model; @Kimura1965; @Lande1976; @Reeve2000; @Bocedi2014] and combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy, respectively, whereby an individual's genotypic values $P_{g}$, $M_{g}$, and $F_{g}$ equal the sum of its 20 alleles for each trait. In overview, each generation proceeds with females paying costs and expressing polyandry, mating, and fertilisation. Offspring inherit a randomly sampled allele from each parent at each locus with no physical linkage. Alleles can then mutate and offspring express inbreeding depression in viability. Immigrants arrive in the population and density regulation limits population growth. We record the population pedigree and directly calculate the coefficient of kinship ($k$) between all potential mates in each generation (defined as the probability that two homologous alleles will be identical-by-descent), allowing individual pre-copulatory and post-copulatory inbreeding strategies to be enacted. Key individual traits, parameter values, and variables are described in Table 1.

@@

Phenotypic values of $P_{p}$, $M_{p}$, and $F_{p}$ each have costs that are assumed to independently increase the probability that a focal female will die before mating (see Discussion for consideration of other costs). The probabilities of pre-mating mortality due to the costs of polyandry ($c_{P}$), pre-copulatory inbreeding strategy ($c_{M}$), and post-copulatory inbreeding strategy ($c_{F}$) are $P_{p} \times c_{P}$, $|M_{p}| \times c_{M}$, and $|F_{p}| \times c_{F}$, respectively. Here $|M_{p}|$ and $|F_{p}|$ are the absolute values of $M_{p}$ and $F_{p}$, respectively. Absolute values are used for applying costs to inbreeding avoidance strategies because both negative and positive $M_{p}$ and $F_{p}$ values could potentially arise and affect the degree of inbreeding, representing inbreeding avoidance and inbreeding preference, respectively (see details of mating and fertilisation strategies below). In contrast, only positive values of $P_{p}$ cause females to be polyandrous (see details of polyandry below). Overall, because generations are non-overlapping, a female's probability of total reproductive failure increases linearly with the phenotypic value of each trait.

@@

The viability of a focal female $i$'s offspring ($\Psi_{\textrm{off}}$) decreases as a log-linear function of her kinship with the sire $j$ of the offspring ($k_{ij}$) and inbreeding depression slope $\beta$,
\begin{equation}
\Psi_{\textrm{off}} = e^{-\beta k_{ij}}
\end{equation}
Here, $\beta$ models the number of haploid lethal equivalents that exist as deleterious recessive alleles in the gametes of $i$ and $j$, and which might be homozygous in offspring and reduce viability. Equation 1 assumes independent allelic effects, generating multiplicative effects on offspring viability [@Morton1956; @Mills1994]. We model inbreeding depression as having an absolute rather than relative effect on offspring viability (i.e., hard rather than soft selection), so that the effect of $\beta$ is consistent across generations and different parameter combinations. We assume that inbreeding always decreases offspring viability (i.e., $\beta > 0$, giving inbreeding depression but no outbreeding depression). Therefore, because $0 \leq k_{ij} \leq 1$, $-\beta \times k_{ij} \leq 0$. Values of $\Psi_{\textrm{off}}$ must therefore be between zero (if $-\beta \times k_{ij}$ is very negative) and one (if $-\beta \times k_{ij} = 0$). We therefore define $\Psi_{\textrm{off}}$ as the probability that an offspring is viable, and sample its realised viability (versus mortality) using a Bernoulli trial. Given our current objectives, we simulate evolution under conditions where inbreeding avoidance is adaptive due to strong inbreeding depression, and not evolution where inbreeding preference is adaptive due to weak or zero inbreeding depression [@Parker1979; @Kokko2006; @Duthie2016a], or given outbreeding depression [@Bateson1983; @Greeff2009]. However, as described above, positive $M_{p}$ and $F_{p}$ values resulting in inbreeding preference are permitted to evolve, and could potentially arise due to mutation or drift.

@@

To address how selection on an initially rare strategy of pre-copulatory or post-copulatory inbreeding avoidance, and resulting evolution, is affected by the other strategy of inbreeding avoidance already being fixed in the population, we first used exploratory simulations to quantify evolution of pre-copulatory inbreeding strategy, and of post-copulatory inbreeding strategy and associated polyandry, in isolation. Then, to test whether pre-copulatory inbreeding avoidance would evolve when adaptive polyandry and post-copulatory inbreeding avoidance were fixed, we initiated $M_{a}$ allele values at zero, but fixed $F_{a}$ allele values at $-10$ and $P_{a}$ allele values at $1$ (i.e., meaning that $F_{p}$ and $P_{p}$ were expressed did not evolve further). Similarly, to test whether post-copulatory inbreeding avoidance would evolve when pre-copulatory inbreeding avoidance was already fixed, we initiated $F_{a}$ and $P_{a}$ allele values at zero but fixed $M_{a}$ allele values at $-10$. Consequently, because females have 10 diploid loci, when $M_{a}$ or $F_{a}$ alleles were fixed at $-10$, outbred females were $51$ times less likely to choose a full brother and $13.5$ times times less likely to choose a first cousin than a non-relative in pre-copulatory and post-copulatory choice, respectively.
